---
title: "Introduction to ppcviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ppcviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
set.seed(2025)
```

# Overview

`ppcviz` implements the novel diagnostic tools from:

> Säilynoja, T., Johnson, A., Martin, R., & Vehtari, A. (2025).
> *Recommendations for visual predictive checks in Bayesian workflow*.
> arXiv:2503.01509.

The core idea is: **treat the visualization itself as a density estimator**
and test whether it faithfully represents the data. A beautiful plot that
misrepresents the distribution is worse than no plot at all.

```{r load}
library(ppcviz)
```

---

## Part 1: Data Property Detection

Before choosing a plot, diagnose your data.

### Discreteness detection

```{r detect-discrete}
# Continuous normal — nothing flagged
detect_discrete(rnorm(300))

# Poisson counts
detect_discrete(rpois(300, lambda = 4))

# Zero-inflated normal
y_mixed <- c(rep(0, 80), rnorm(220, mean = 2))
detect_discrete(y_mixed)
```

### Boundary detection

```{r detect-bounds}
# Unbounded
detect_bounds(rnorm(300))

# Non-negative
detect_bounds(rexp(300))

# Proportions [0, 1]
detect_bounds(rbeta(300, 2, 5))
```

---

## Part 2: Visualization PIT Values

The *visualization PIT* (Probability Integral Transform) quantifies how well
a chosen plot type represents the empirical distribution.  If the PIT values
are uniform on \([0,1]\), the visualization is faithful; non-uniformity
reveals where the plot is misleading.

### KDE PIT

```{r pit-kde}
y <- rnorm(300)
pit <- pit_kde(y)
hist(pit, breaks = 25, main = "KDE PIT — normal data (should be ~uniform)",
     xlab = "PIT", col = "#DCE0F5", border = "white")
```

#### Bounded data

Without a boundary correction, a KDE on \([0,1]\) data leaks density outside
the valid range, causing systematic non-uniformity:

```{r pit-kde-bounded}
y_beta <- rbeta(300, 2, 5)

# Without bounds — PIT will pile up near 0 and 1
pit_bad <- pit_kde(y_beta)
hist(pit_bad, breaks = 25, main = "KDE PIT — Beta data WITHOUT bounds",
     xlab = "PIT", col = "#FFD0D0", border = "white")

# With reflection correction — should be ~uniform
pit_ok <- pit_kde(y_beta, bounds = c(0, 1))
hist(pit_ok, breaks = 25, main = "KDE PIT — Beta data WITH bounds = c(0,1)",
     xlab = "PIT", col = "#DCE0F5", border = "white")
```

### Histogram PIT

```{r pit-histogram}
pit_h <- pit_histogram(y)
hist(pit_h, breaks = 25, main = "Histogram PIT — normal data",
     xlab = "PIT", col = "#DCE0F5", border = "white")
```

### Quantile dot plot PIT

```{r pit-qdotplot}
pit_q <- pit_qdotplot(y, n_quantiles = 50L)
hist(pit_q, breaks = 25, main = "Quantile dot plot PIT — normal data",
     xlab = "PIT", col = "#DCE0F5", border = "white")
```

---

## Part 3: Goodness-of-Fit for Visualizations

`viz_gof()` tests whether the PIT values are consistent with uniformity using
*simultaneous* ECDF confidence bands (Säilynoja, Bürkner & Vehtari 2022).

```{r viz-gof-pass}
# Normal data with KDE — should pass
result_pass <- viz_gof(pit_kde(y), n_sim = 500)
print(result_pass)
plot(result_pass)
```

```{r viz-gof-fail}
# Beta data without boundary correction — may fail
result_fail <- viz_gof(pit_kde(y_beta), n_sim = 500)
print(result_fail)
plot(result_fail)
```

### Convenience wrapper

`check_viz()` combines PIT computation and the GoF test and prints a
human-readable verdict:

```{r check-viz}
check_viz(y, method = "kde")
check_viz(y_beta, method = "kde")
check_viz(y_beta, method = "kde", bounds = c(0, 1))
```

---

## Part 4: PAVA Calibration Plots

These plots are the paper's main contribution for **binary and discrete
outcomes**. They do not exist in `bayesplot`.

### Binary calibration

```{r calibration-binary}
n <- 200
true_p <- plogis(rnorm(n))
y_bin  <- rbinom(n, 1, true_p)
yrep   <- do.call(rbind, lapply(seq_len(100), function(i) rbinom(n, 1, true_p)))

# Well-calibrated predictions — should be near the diagonal
ppc_calibration(y_bin, yrep, n_boot = 200)
```

```{r calibration-residual}
ppc_calibration_residual(y_bin, yrep, n_boot = 200)
```

### Discrete / ordinal calibration

```{r calibration-discrete}
y3    <- sample(0:2, n, replace = TRUE)
yrep3 <- do.call(rbind, lapply(seq_len(100), function(i) sample(0:2, n, replace = TRUE)))
ppc_calibration_discrete(y3, yrep3, n_boot = 100)
```

---

## Part 5: Automatic Recommender

`recommend_ppc()` combines all the detection and testing logic to suggest the
most appropriate `bayesplot` (or `ppcviz`) function for your data:

```{r recommend-continuous}
recommend_ppc(rnorm(300))
```

```{r recommend-bounded}
recommend_ppc(rexp(300))
```

```{r recommend-proportion}
recommend_ppc(rbeta(300, 2, 5))
```

```{r recommend-discrete}
recommend_ppc(rpois(300, lambda = 3))
```

```{r recommend-mixed}
recommend_ppc(c(rep(0, 70), rnorm(230, mean = 2)))
```

```{r recommend-binary}
recommend_ppc(sample(0:1, 300, replace = TRUE))
```

---

## Citation

If you use `ppcviz` in a publication, please cite the original paper:

```
Säilynoja, T., Johnson, A., Martin, R., & Vehtari, A. (2025).
Recommendations for visual predictive checks in Bayesian workflow.
arXiv:2503.01509.
```
